---
title: "EDS 240 HW3: Visualizing FEMA NRI x ACS Data"
author: "Nathalie Bonnet"
date: 2026-02-11
format: html
message: FALSE
warning: FALSE
---
# Setup

Reading in data
```{r}
# Read in libraries
library(tidyverse)
library(here)
library(janitor)
library(stringr)
```


```{r}
# # #....Step 1a: see all available ACS variables + descriptions.....
# acs_vars <- tidycensus::load_variables(year = 2023,
#                                       dataset = "acs1")
#
# # #..............Step 1b: import race & ethnicity data.............
# race_ethnicity <- tidycensus::get_acs(
# geography = "county",
#   survey = "acs1",
#    # NOTE: you may not end up using all these variables
#    variables = c("B01003_001", "B02001_002", "B02001_003",
#                  "B02001_004", "B02001_005", "B02001_006",
#                  "B02001_007", "B02001_008", "B03002_012",
#                  "B03002_002"),
#    state = "CA",
#    year = 2023) |>
# #   # join variable descriptions (so we know what's what!)
#    dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name))
#
# # #.................Step 2: write ACS data to file.................
#  readr::write_csv(race_ethnicity, here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- readr::read_csv(here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

# Read in NRI and filter for only California
nri <- read_csv(here(
  "data",
  "National_Risk_Index_Counties_807384124455672111.csv"
)) %>%
  clean_names() %>%
  filter(state_name %in% "California")
```

# Building vizualization
goal: Create a data viz that helps to answer the question, How does climate hazard risk exposure vary across racial / ethnic groups in California?

## Data cleaning and join
```{r}
# Clean new ACS data
re_df <- race_ethnicity %>%
  # convert column names to lower snake case
  clean_names() %>%
  # Split column on comma
  separate_wider_delim(
    cols = name,
    delim = ",",
    names = c("county_name", "state")
  ) %>%
  # remove all text before !!
  mutate(label = str_remove(label, pattern = ".*!!")) %>%
  # Filter out random non hispanic and latino rows
  filter(!label %in% "Not Hispanic or Latino:") %>%
  # Remove colons and the word alone from race/ethnicity label
  mutate(label = str_remove(label, ":")) %>%
  mutate(label = str_remove(label, "alone")) %>%
  # Remove county part of county name and any whitespace
  mutate(county_name = str_remove(county_name, "County")) %>%
  mutate(county_name = str_squish(county_name))


# Join dataframe on county
nri_acs <- full_join(x = re_df, y = nri, by = "county_name") %>%
  # Select only desired columns
  select(county_name,
         estimate,
         label,
         national_risk_index_rating_composite)
```

# Preliminary data wrangling and viz
```{r}
# Create new data frame with percentages of each group represented by NRI risk rating
perc_risk_cat <- nri_acs %>%
  # Grouping by race/ethnicity and rating category
  group_by(label, national_risk_index_rating_composite) %>%
  # Create new column with the total number of people in those categories
  summarize(total_pop = sum(estimate)) %>%
  # Drop NAs
  drop_na(label) %>%
  # Group again by group identity
  group_by(label) %>%
  # New column with the percentage of each group in each category
  mutate(perc = total_pop / sum(total_pop)) %>%
  # Remove "total" as a group
  filter(!label %in% "Total")


# Reorder based on amount of people at high risk
label_order <- perc_risk_cat %>%
  filter(national_risk_index_rating_composite == "Very High") %>%
  arrange(desc(perc)) %>%
  pull(label)

# Overwrite dataframe with levels
perc_risk_cat <- perc_risk_cat %>%
  mutate(
    # Order lable by descending high risk percent
    label = factor(label, levels = label_order),
    # reorder ratings so they show up high to low on graph
    national_risk_index_rating_composite = factor(
      national_risk_index_rating_composite,
      levels = c(
        "Relatively Low",
        "Relatively Moderate",
        "Relatively High",
        "Very High"
      )
    )
  )
```

```{r}
ggplot(data = perc_risk_cat,
  aes(
    x = label,
    y = perc,
    # Group and color by rating
    group = national_risk_index_rating_composite,
    fill = national_risk_index_rating_composite
  )
) +
  # Stacked cols chart
  geom_col() +
  # Flip race and ethnicity labels to y
  coord_flip() +
  # Remove padding around axes
  scale_y_continuous(expand = c(0, 0))+
  scale_x_discrete(expand = c(0, 0))+
  # Fill discrete colors
  scale_fill_viridis_d(
    option = "rocket",
    # Reverse for darker = higher
    direction = -1,
    # Shrink color scale for visuals
    begin = 0.2,
    end = 0.8
  ) +
  # Minimal theme
  theme_minimal() +
  # Edit legend
  guides(fill = guide_legend(
    # Legend title
    title = "Risk Rating",
    # Place title on bottom of legend
    title.position = "bottom",
    # Put legend on the bottom of the chart
    position = "bottom",
    # Reverse order for readability
    reverse = TRUE
  )) +
   # Add plot labels, caption, and alt text
  labs(
    title = "National Risk Index (NRI) Rating for Race/Ethnic Groups in California",
    subtitle = "all groups are at high risk for impact by natural hazards",
    x = "Race/Ethnicity group", 
    caption = "Data: FEMA Resilience Analysis and Planning Tool (RAPT) National Risk Index (NRI) (2025 Release), &  U.S. Census Bureau’s American Community Survey (ACS) (2023).",
    alt = "Proportional stacked bar chart showing the proportion of race/ethnic groups in California in eachmrelative National Risk Index (NRI) rating category (Very High to Very Low). The groups being represented are White, Black or African American, American Indian and Alaska Native, Asian, Native Hawaiian and Other Pacific Islander, Some Other Race, Two or More Races, and Hispanic or Latino. 50 to over 0.75 percent of each group experienced Very High risk of impact by natural hazards, with the group at most high risk being Asian. The group with the greatest proportion of Moderate Risk individuals was White."
  )+
  # Adjust theme elements
  theme(
    axis.title.x = element_blank(), 
    # Adjust title position
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(hjust = 0.5)
  )

```

# Discussion

1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?

In this visualization, I was interested in 

2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?

3. Summarize your main finding in no more than two sentences.

4. What modifications did you make to this visualization to make it more easily readable?

5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.


